Data Provenance for Concurrent Program Debugging and Analysis in Production Systems

### 2013-11-12 paper.tex
edits to intro

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to the simplicity of
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like causing a crash or degrading performance.

@@

Debugging a concurrent program requires understanding how threads interact and
{\em data provenance information} furnishes that understanding. A value's
provenance is a description of the reason that it is stored in a variable.
Provenance can take on many forms, for example, a value's provenance could be
the instruction that wrote it or the function that returned it.  Provenance
information helps answer the question: "Why does this variable contain this
value?", which is central to debugging.

@@

Collecting debugging information, like data provenance, from production systems
is particularly important for concurrent software.  Concurrency bugs manifest
as failures rarely.  When a production system fails and a programmer receives a
bug report, reproducing the reported failure to collect provenance information
may be difficult or impossible.  It is essential to collect as much information
as performance margins permit from production, to reduce the need for
time-consuming failure reproduction.

Prior work has previously shown that some limited forms of provenence help with
debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort tracked
code that wrote unusable values into variables~\cite{badapples}.  That work
showed that revealing the connection between an unusable value's origin and a
failure it causes is helpful during debugging.  Other work showed that
programmers debug more effectively when they can ask "why" questions about
values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Addressing
concurrency is important because a value could have originated in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult.

In this work we design a novel form of in-production provenance tracking for
concurrent programs with three goals.  (1) To collect a general form of
provenance information, not limited to only certain values.  (2) To provide
provenance information that helps understand how threads interact, which is the
key to understanding concurrency bugs.  (3) To work with overheads low enough
for use in production (ideally, less than 10\%).

@@

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
potentially shared memory location when a thread accesses such a location.  If
the accessing thread is different from the thread that last wrote the location,
the access signifies {\em communication} between the threads.  When
communication occurs, \ctraps delivers a software {\em trap} to the accessing
thread, giving it information about how communication occurred.  The trap can be
handled by an analysis-specific trap handler.  Trap handlers implement {\em
\ctraps applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}).  By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases.  \ctraps applications can perform arbitrary concurrency analyses,
but at their own discretion and cost.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.

>>>>

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to 
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like causing a crash or degrading performance.

@@

A value's provenance is a description of the reason that it is stored in a
variable. Provenance information helps answer the question: "Why does this
variable contain this value?", which is central to debugging. Debugging a
concurrent program requires understanding how threads interact, and since
communication happens via reads and writes to shared variables, {\em data
provenance information is essential}. Provenance can take on many forms, for
example, a value's provenance could be the instruction that wrote it, the
function that returned it, and in the context of concurrent programs, the
thread that produced it.

@@

Collecting debugging information, like data provenance, from production
systems is particularly important for concurrent software since bugs
manifest as failures rarely. When a production system fails and a programmer
receives a bug report, reproducing the reported failure to collect provenance
information may be difficult or impossible. It is essential to collect as much
information as performance margins permit from production, to reduce the need
for time-consuming failure reproduction.

Prior work has previously shown that some limited forms of provenance help
with debugging~\cite{badapples, whylineicse, tipslicingsurvey}. One effort
tracked code that wrote unusable values into variables~\cite{badapples}. That
work showed that revealing the connection between an unusable value's origin
and a failure it causes is helpful during debugging. Other work showed that
programmers debug more effectively when they can ask ``why'' questions about
values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance
tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive. Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples}{\bf TODO: Brandon, check this reference, since you just cited it in the previous sentence.} impose overheads too high for production for
C and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems. Addressing
concurrency is important because a value could have originated in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult. In summary, the challenges in using provenance to debug concurrent programs are: (1) collecting the right information to characterize the concurrent execution behavior; and (2) making it efficient enough for general programs running in production settings.

In this work we address the challenges above and develop a novel form of
in-production provenance tracking for concurrent programs. We have three
goals. (1) To collect a general form of provenance information, not limited to
only certain values. (2) To provide provenance information that helps
understand how threads interact, which is the key to understanding concurrency
bugs. (3) To work with overheads low enough for use in production (ideally,
less than 10\%).

@@

Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
(Section~\ref{sec:ctraps}). \ctraps can be used to implement several dynamic
analysis tailored to concurrent programs, like the ones discussed above.
\ctraps works by checking the last writer slice of each potentially shared
memory location when a thread accesses such a location. If the accessing
thread is different from the thread that last wrote the location, the access
signifies {\em communication} between the threads. When communication occurs,
\ctraps delivers a software {\em trap} to the accessing thread, giving it
information about how communication occurred. The trap can be handled by an
analysis-specific trap handler. Trap handlers implement {\em \ctraps
applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}). By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases. 
%\ctraps applications can perform arbitrary concurrency analyses,
%but at their own discretion and cost. 
As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.
